---
title: "Bayesian Econometrics: Homework 1"
author: "Jerin Varghese"
date: "February 1, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

Download and install R and R-Studio from https://cran.r-project.org and
https://www.rstudio.com/products/rstudio/download/

## Problem 2
Use R to calculate the following:

1. Create a vector called x containing the number 2 5 8 12 16
```{r}
x <- c(2,5,8,12,16)
```

2. Create a vector called y containing the number from 5 to 9
```{r}
y <- c(5:9)
```

3. Calculate z = x-y
```{r}
z <- x-y
```
```{r, echo=FALSE}
z
```

## Problem 3
1. Create a vector x of 100 values starting at 2 and increasing by 3 each time
```{r}
x <- seq(from=2,by=3,length.out=100)
```

2. Extract the values at positions 5,10,15 and 20 in the vector of values from x
```{r}
x[seq(5,20,5)]
```


3. Extract the values at positions 10 to 20.
```{r}
x[10:20]
```

4. Create 25 by 4 matrix x3 from x (first 4 elements go to the first row, next 4 elements
go to 2nd row, etc.)
```{r}
x3 <- matrix(data=x, nrow=25, ncol=4)
```

5. Create 25 by 2 matrix from 2nd and 3rd column of matrix x3.
```{r}
matrix(data=x3[,2:3], nrow=25, ncol=2)
```


## Problem 4
1. Save a tap delimited file, called sample file.txt in a directory and set your working
directory to where your data is stored. Then read the file into a new structure using
read.delim command and check (print) the row 11.
```{r}
working_dir <- "C:/Users/jerin/OneDrive/Documents/JHU/Bayesian Econometrics/Homeworks"
setwd(working_dir)
sample_file <- read.delim('sample_file.txt')
sample_file[11,]
```

2. Calculate the mean of the column named Length.
```{r}
mean(sample_file$Length)
```

3. Find out how many rows in sample file have a Length which is < 65.
```{r}
nrow(sample_file[sample_file$Length > 65,])
```

4. Print the values of Length less than 65.
```{r}
sample_file[sample_file$Length > 65,"Length"]
```


## Problem 5.
1. Generate a sequence of x variable from 0 to 10 by 0.5 increment.
```{r}
x <- seq(0, 10, 0.5)
```

2. Generate yi = 2 + 3*xi + ei where ei ~ i.i.d.N(0, 3^2).
```{r}
y <- 2 + 3*x + rnorm(length(x),mean=0,sd=3)
```

3. Plot x and y.
```{r}
plot(x,y)
```

4. Fit a linear regression model for y on x using command lm
```{r}
lm.fit <- lm(y~x)
```

5. Draw the fitted regression line.
```{r}
plot(x,y) + abline(lm.fit, col="red")
```


## Problem 6.
1. Draw a sample of 100 observations from gamma distribution with shape parameter =1
and rate parameter=4.
```{r}
set.seed(42)
gsample <- rgamma(100, shape=1, rate=4)
```

2. calculate the mean and compare with the theoretical mean.
```{r}
# sample mean is .246, which is very close to theoretical mean of 1/4
mean(gsample)
```

3. Draw histogram and density of the sample
```{r}
hist(gsample, prob=TRUE)
```


## Problem 7
### Generate 1000 observations of an AR process and draw graphs:
1. generate a standard normal random variable for the initial value x1.
```{r}
set.seed(42)
n <- 1000
white_noise <- rnorm(n)
# initialize first value
ar1 <- c(white_noise[1])
```

2. for xi from i=2 to 1000, xi = .09*x[i-1] + ei where ei ~ i.i.d.N(0, 1)
```{r}
for (i in 2:n){
  ar1[i] <- .09*ar1[i-1] + white_noise[i]
}
```

3. plot x and cumulative sum of x (use command cusum) over i
```{r}
plot(ar1, cumsum(ar1))
```



